Practical Optional Types for Clojure

### 2015-02-21 formal-model.tex
remove class table lookup figure from main paper

Syntax is given in figure~\ref{figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals, do (sequencing) expressions and let expressions.
We model throwable errors without a catching mechanism. All binding forms introduce fresh
variables.

@@

Typing rules are in figure~\ref{figure:typingrules}.
T-Error neither returns logically true or false as it does not return normally.
T-Do uses the propositions learnt in the first expression to type check the second.
Other than compensating for the extra logically false value \nil{}, the rest are standard.

Figure~\ref{figure:standardopsem} contains the operational semantics, which are standard
except for \nil{} being false.

Subtyping (figure~\ref{figure:subtyping}) says singleton types are instances of their respective classes
and \Object is transitively a supertype to everything except {\Nil{}},
so \Top{} = {\Union{\Nil}{\Object}}.

@@

                &\mbox{Constants}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                      \alt {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}
                &\mbox{Types} \\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}\\
  \openvsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\label{figure:termsyntax}
\end{figure*}

@@

    {\TSubsume}
  \end{mathpar}
  \caption{Typing rules}
  \label{figure:typingrules}
\end{figure*}

@@

    \BDelta{}
  \end{mathpar}
  \caption{Operational Semantics}
  \label{figure:standardopsem}
\end{figure*}

\begin{figure*}
  \footnotesize
\begin{mathpar}
\standardsubtyping{}
\end{mathpar}
\caption{Subtyping rules}
\label{figure:subtyping}
\end{figure*}

@@

With an elegant extension in T-Do (figure~\ref{figure:typingrules}) we can follow this reasoning
to infer \mintinline{clojure}{(inc x)} is safe. The main insight is we can assume the
disjunction of the propositions of the first \doliteral{} argument while checking the second.

@@

One interesting Clojure function we modeled is \isaliteral.
It takes two arguments, and returns \true{} if they are the same (equality mode)
or if they are both classes and the left is a subtype of the right (subclassing mode),
otherwise returns \false{}, as described in figure~\ref{figure:isasemantics}.

@@

We encode this behaviour in figure~\ref{figure:isatyping}. 
\isacompareliteral{} is the interesting part,
which takes the object of the first argument 
and the type of the second argument, and returns the then and else propositions
of the expression.

@@

    \BIsA{}
  \end{mathpar}
  \caption{\isaliteral{} semantics}
  \label{figure:isasemantics}
\end{figure}

@@

    \TIsA{}
  \end{mathpar}
  \caption{\isaliteral{} typing rules}
  \label{figure:isatyping}
\end{figure}

@@

 \disptablesyntax{} \\
\end{altgrammar}
$$
\caption{Multimethod Syntax}
\label{figure:mmsyntax}
\end{figure*}

@@

    \TDefMethod{}
  \end{mathpar}
  \caption{Multimethod typing rules}
  \label{figure:mmtyping}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \Multisubtyping{}
  \end{mathpar}
  \caption{Multimethod subtyping rules}
  \label{figure:mmsubtype}
\end{figure}

@@

    \BBetaMulti{}
  \end{mathpar}
  \caption{Multimethod semantics}
  \label{figure:mmsemantics}
\end{figure}

@@

To ease presentation, we present \emph{immutable}
multimethods (figure~\ref{figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. The above is now written

@@

Multimethod semantics are in figure~\ref{figure:mmsemantics}.
B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
B-DefMethod returns a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
which \getmethodliteral{} uses to choose the method to invoke.

The typing rules (figure~\ref{figure:mmtyping}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
where \s{} is the interface type for beta reduction, and \t{} is the type for
the dispatch function. T-DefMulti uses the provided interface type and dispatch function
to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
that must be true if this method is being invoked. The example can be type checked without
changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
method bodies. T-App handles beta reduction because multimethods are subtypes of functions
(figure~\ref{figure:mmsubtype}).

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\emptymap{}}
    &\mbox{Values} \\
    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
    &\mbox{Types} \\
    \auxhmapsyntax{}\\
  \end{altgrammar}
  $$
  \caption{HMap Terms and Types}
  \label{figure:hmapsyntax}
\end{figure}

@@

    {\BGetMissing}
  \end{mathpar}
  \caption{HMap Semantics}
  \label{figure:hmapsem}
\end{figure}

@@

    {\TAssoc}
  \end{mathpar}
  \caption{HMap Typing Rules}
  \label{figure:hmaptyping}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \HMapsubtyping{}
  \end{mathpar}
  \caption{HMap subtyping rules}
  \label{figure:hmapsubtype}
\end{figure}

It is common in Clojure to use plain immutable hash maps to represent ad-hoc
objects with a known set of keyword entries. We use \HMapliteral{}
types (figure~\ref{figure:hmapsyntax}) to track known present, known absent and unknown keyword entries.

@@

Figure~\ref{figure:hmapsem} shows the semantics for HMaps, where B-Get returns the associated
entry if it is present and B-GetMissing returns \nil{} otherwise. B-Assoc returns a new extended
map.

Typing rules are given in figure~\ref{figure:hmaptyping}. T-GetHMap handles the simple
case where we know a lookup will succeed and use B-Get. In T-GetHMapAbsent we know B-GetMissing will give
a \Nil{} return type.
T-GetHMapPartialDefault does not know if the key is present or absent and has a \Top{} return type.
T-Assoc extends the known keyword entries, overwriting previous entries with the same key.

S-HMap (figure~\ref{figure:hmapsubtype}) says that HMap subtypes must have all present entries to also be
subtypes and also cover the absent keys.
Complete maps are subtypes of partial maps by S-HMapP.

@@

\begin{figure*}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
            \alt  \localhintsyntax{} \alt \lethintsyntax{}
    &\mbox{Expressions}
  \end{altgrammar}
  $$
  \caption{Java Interoperability Terms and Types}
  \label{figure:reflectivesyntax}
\end{figure*}

@@

We extend the external language with Java field lookups and calls to
methods and constructors (figure~\ref{figure:nonreflectivesyntax}).
There are no typing or evaluation rules for these syntaxes because
Typed Clojure must be convinced that Java interactions are \emph{non-reflective},
ie., are unambiguously resolved at compile-time. Programmers use
type hinted expressions (figure~\ref{figure:nonreflectivesyntax}) 
to help the compiler eliminate reflection.

@@

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \caption{Internal Language Reflection Resolution Extensions}
  \label{figure:nonreflectivesyntax}
\end{figure*}

@@

    \RLetHint{}
  \end{mathpar}
\caption{Reflection Elimination}
\label{figure:rewrite}
\end{figure*}

@@

    \TALet{}
  \end{mathpar}
\caption{Type Hint Inference}
\label{figure:hintinfer}
\end{figure*}

@@

    {\TMethodStatic}
  \end{mathpar}
  \caption{Java Interoperability Typing Rules}
  \label{figure:javatyping}
\end{figure}

@@

core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes resolved
reflection, however to tell the full story we include a rewrite relation 
in figure~\ref{figure:rewrite} to eliminate reflection when sufficient type hints are provided.
Figure~\ref{figure:hintinfer} shows the interesting cases in type hint inference.

We now have two languages: an internal language that may contain non-reflective Java calls (figure 
  \ref{figure:nonreflectivesyntax}) 
  and an external language that resembles Clojure, where we must rely on type-hints to resolve
  reflective Java calls (figure~\ref{figure:reflectivesyntax}).

@@

We then extract these type hints in T-FieldStatic for typing checking 
(figure~\ref{figure:javatyping}), which converts type hints
with \javatotcliteral{} to obtain a \emph{non-nilable} type to type check the target expression (which prevents
dereferencing nil) and \javatotcnilliteral{} to get a \emph{nilable} return type (because fields can be nil).

@@

\input{constant}
\input{update}
\input{convert-types-figure}

>>>>

Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals, do (sequencing) expressions and let expressions.
We model throwable errors without a catching mechanism. All binding forms introduce fresh
variables.

@@

Typing rules are in figure~\ref{main:figure:typingrules}.
T-Error neither returns logically true or false as it does not return normally.
T-Do uses the propositions learnt in the first expression to type check the second.
Other than compensating for the extra logically false value \nil{}, the rest are standard.

Figure~\ref{main:figure:standardopsem} contains the operational semantics, which are standard
except for \nil{} being false.

Subtyping (figure~\ref{main:figure:subtyping}) says singleton types are instances of their respective classes
and \Object is transitively a supertype to everything except {\Nil{}},
so \Top{} = {\Union{\Nil}{\Object}}.

@@

                &\mbox{Constants}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                      \alt {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}
                &\mbox{Types} \\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}\\
  \openvsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\label{main:figure:termsyntax}
\end{figure*}

@@

    {\TSubsume}
  \end{mathpar}
  \caption{Typing rules}
  \label{main:figure:typingrules}
\end{figure*}

@@

    \BDelta{}
  \end{mathpar}
  \caption{Operational Semantics}
  \label{main:figure:standardopsem}
\end{figure*}

\begin{figure*}
  \footnotesize
\begin{mathpar}
\standardsubtyping{}
\end{mathpar}
\caption{Subtyping rules}
\label{main:figure:subtyping}
\end{figure*}

@@

With an elegant extension in T-Do (figure~\ref{main:figure:typingrules}) we can follow this reasoning
to infer \mintinline{clojure}{(inc x)} is safe. The main insight is we can assume the
disjunction of the propositions of the first \doliteral{} argument while checking the second.

@@

One interesting Clojure function we modeled is \isaliteral.
It takes two arguments, and returns \true{} if they are the same (equality mode)
or if they are both classes and the left is a subtype of the right (subclassing mode),
otherwise returns \false{}, as described in figure~\ref{main:figure:isasemantics}.

@@

We encode this behaviour in figure~\ref{main:figure:isatyping}. 
\isacompareliteral{} is the interesting part,
which takes the object of the first argument 
and the type of the second argument, and returns the then and else propositions
of the expression.

@@

    \BIsA{}
  \end{mathpar}
  \caption{\isaliteral{} semantics}
  \label{main:figure:isasemantics}
\end{figure}

@@

    \TIsA{}
  \end{mathpar}
  \caption{\isaliteral{} typing rules}
  \label{main:figure:isatyping}
\end{figure}

@@

 \disptablesyntax{} \\
\end{altgrammar}
$$
\caption{Multimethod Syntax}
\label{main:figure:mmsyntax}
\end{figure*}

@@

    \TDefMethod{}
  \end{mathpar}
  \caption{Multimethod typing rules}
  \label{main:figure:mmtyping}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \Multisubtyping{}
  \end{mathpar}
  \caption{Multimethod subtyping rules}
  \label{main:figure:mmsubtype}
\end{figure}

@@

    \BBetaMulti{}
  \end{mathpar}
  \caption{Multimethod semantics}
  \label{main:figure:mmsemantics}
\end{figure}

@@

To ease presentation, we present \emph{immutable}
multimethods (figure~\ref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. The above is now written

@@

Multimethod semantics are in figure~\ref{main:figure:mmsemantics}.
B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
B-DefMethod returns a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
which \getmethodliteral{} uses to choose the method to invoke.

The typing rules (figure~\ref{main:figure:mmtyping}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
where \s{} is the interface type for beta reduction, and \t{} is the type for
the dispatch function. T-DefMulti uses the provided interface type and dispatch function
to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
that must be true if this method is being invoked. The example can be type checked without
changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
method bodies. T-App handles beta reduction because multimethods are subtypes of functions
(figure~\ref{main:figure:mmsubtype}).

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\emptymap{}}
    &\mbox{Values} \\
    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
    &\mbox{Types} \\
    \auxhmapsyntax{}\\
  \end{altgrammar}
  $$
  \caption{HMap Terms and Types}
  \label{main:figure:hmapsyntax}
\end{figure}

@@

    {\BGetMissing}
  \end{mathpar}
  \caption{HMap Semantics}
  \label{main:figure:hmapsem}
\end{figure}

@@

    {\TAssoc}
  \end{mathpar}
  \caption{HMap Typing Rules}
  \label{main:figure:hmaptyping}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \HMapsubtyping{}
  \end{mathpar}
  \caption{HMap subtyping rules}
  \label{main:figure:hmapsubtype}
\end{figure}

It is common in Clojure to use plain immutable hash maps to represent ad-hoc
objects with a known set of keyword entries. We use \HMapliteral{}
types (figure~\ref{main:figure:hmapsyntax}) to track known present, known absent and unknown keyword entries.

@@

Figure~\ref{main:figure:hmapsem} shows the semantics for HMaps, where B-Get returns the associated
entry if it is present and B-GetMissing returns \nil{} otherwise. B-Assoc returns a new extended
map.

Typing rules are given in figure~\ref{main:figure:hmaptyping}. T-GetHMap handles the simple
case where we know a lookup will succeed and use B-Get. In T-GetHMapAbsent we know B-GetMissing will give
a \Nil{} return type.
T-GetHMapPartialDefault does not know if the key is present or absent and has a \Top{} return type.
T-Assoc extends the known keyword entries, overwriting previous entries with the same key.

S-HMap (figure~\ref{main:figure:hmapsubtype}) says that HMap subtypes must have all present entries to also be
subtypes and also cover the absent keys.
Complete maps are subtypes of partial maps by S-HMapP.

@@

\begin{figure*}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
            \alt  \localhintsyntax{} \alt \lethintsyntax{}
    &\mbox{Expressions}
  \end{altgrammar}
  $$
  \caption{Java Interoperability Terms and Types}
  \label{main:figure:reflectivesyntax}
\end{figure*}

@@

We extend the external language with Java field lookups and calls to
methods and constructors (figure~\ref{main:figure:nonreflectivesyntax}).
There are no typing or evaluation rules for these syntaxes because
Typed Clojure must be convinced that Java interactions are \emph{non-reflective},
ie., are unambiguously resolved at compile-time. Programmers use
type hinted expressions (figure~\ref{main:figure:nonreflectivesyntax}) 
to help the compiler eliminate reflection.

@@

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \caption{Internal Language Reflection Resolution Extensions}
  \label{main:figure:nonreflectivesyntax}
\end{figure*}

@@

    \RLetHint{}
  \end{mathpar}
\caption{Reflection Elimination (select rules, figure~\ref{appendix:figure:rewrite} for full rules)}
\label{main:figure:rewrite}
\end{figure*}

@@

    \TALet{}
  \end{mathpar}
\caption{Type Hint Inference (select rules, figure~\ref{appendix:figure:hintinfer} for full rules)}
\label{main:figure:hintinfer}
\end{figure*}

@@

    {\TMethodStatic}
  \end{mathpar}
  \caption{Java Interoperability Typing Rules}
  \label{main:figure:javatyping}
\end{figure}

@@

core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes resolved
reflection, however to tell the full story we include a rewrite relation 
in figure~\ref{main:figure:rewrite} to eliminate reflection when sufficient type hints are provided.
Figure~\ref{main:figure:hintinfer} shows the interesting cases in type hint inference.

We now have two languages: an internal language that may contain non-reflective Java calls (figure 
  \ref{main:figure:nonreflectivesyntax}) 
  and an external language that resembles Clojure, where we must rely on type-hints to resolve
  reflective Java calls (figure~\ref{main:figure:reflectivesyntax}).

@@

We then extract these type hints in T-FieldStatic for typing checking 
(figure~\ref{main:figure:javatyping}), which converts type hints
with \javatotcliteral{} to obtain a \emph{non-nilable} type to type check the target expression (which prevents
dereferencing nil) and \javatotcnilliteral{} to get a \emph{nilable} return type (because fields can be nil).

@@

\input{constant}
\input{update}

{\convertjavatypefigure{figure}}
